---
title: "Exam 2016"
author: "Carey Hedges (751546)"
output: github_document
---

#Question 1 - Body Temperature and Resting Heart Rate

Temperature is interval continuous data and heart rate is discrete numerical data.  The gender data is nominal data.

The hypotheses are:

Ho: Relations between body temperature and heart rate are sex independent.
H1: Relations between body temperature and heart rate are sex dependent.

A correlation and linear regression analysis should be used to establish if the temperature and heart rate are related in both male and female populations independently.  

The assumptions underlying a linear regression are:

1) A trend exists between heart rate and body temperature.
2) Observations are independent of one another.
3) Measurements of the independent variable have been made correctly.
4) Residuals are normally distributed and are homoskedastic (same variance).

Assumptions underlying Pearson's correlation analyses (as used here) are:

1) Data should be interval or ratio data.
2) Data should have a linear relationship.
3) Data should not have any outliers or pivot points.
4) Variables should both be approximately normally distributed.

These tests shouyld indicate if there is a relation between temperature and resting heart rate in different genders. 

```{r question 1, echo = TRUE, fig.height=8, fig.width=8, fig.align='center', fig.path='./figures/', dev = 'png'}


library(knitr)
library(dplyr)

#Instruct R Studio to read the excel spreadsheet

 A <- read.csv('C:/Users/Carey Hedges/Desktop/Stats - Exam/question1.csv')
 A
 
#Gain a holistic overview of the dataset
 
 summary(A)
 head(A)
 colnames(A)
 tail(A)
 
#Isolate male data 
 
AM <- A [,-3]
AM

#Isolate female data

AF <- A [,-2]
AF

#Plot data to visualise, and run linear model to assess if the data are correlated.

#Plot of male data

plot(AM$body_temperature, AM$male, 
     main = "Heart Rate versus Body Temperature for Male Subjects",
     xlab = "Body Temperature",
     ylab = "Resting Heart Rate")
LAM <- lm(AM)
summary(LAM)

#As the male data has no linear relationship, a correlation cannot be performed.

#Plot of female data

plot(AF$body_temperature, AF$female, 
     main = "Heart Rate versus Body Temperature for Female Subjects",
     xlab = "Body Temperature",
     ylab = "Resting Heart Rate")
LAF <- lm(AF)
summary(LAF)

CAF <- cor.test(AF$body_temperature, AF$female, method = "pearson")
CAF

#You are required to check that the data is homoskedatic.  Male data is not linear so only female data is used here.

qqnorm(LAF$residuals) 
qqline(LAF$residuals)


```

Data analysis indicates that there is no linear relationship between the heart rate and body temperature of male subjects, b|= 0.1244, t(63)=1.521, p = 0.1333.  There is a weak positive linear relationship (r=0.2840, p=0.02186) between the body temperature and heart rate in females b|= 0.1445, t(63)=2.351, p=0.0219, R^2^ = 0.08066 (adjusted R^2^ = 0.06607), F(1, 63) = 5.528, p=0.02186.  The data from the female population are homoskedastic and do not vialate the basic assumptions of linear regression analyses.  This validates the linear model in the female group.

These data indicate that body temperature and heart rate are sex dependent and females with higher resting heart rates will have higher oral body temperature. This inference would be sound for all body temperatures between 35.70 degrees celcius and 38.20 degrees celcius.

#Question 2 - Intoxicated Ataxia

The all of the data variables (that is, left hand, right hand, males, female, line left, line right) are categorical nominal data.

The hypotheses are: 

Ho: Handedness does not affect body position and limb favouring in intoxicated subjects.
H1: Handedness does affect body position and limb favouring in intoxicated subjects.

A Fisher's exact test will be used to evaluate if handedness affects body position in intoxicated subjects. 

The assumptions underlying a Fischer's exact test are :

1. Values are independent of one another 
2. Sampling is random
3. Observed frequencies are approximated by normal distribution
4. Expected values should be 5 or greater in all cells 

```{r question 2, echo = TRUE, fig.height=8, fig.width=8, fig.align='center', fig.path='./figures/', dev = "png"}

library(knitr)
library(dplyr)

#Get R to read the excel spreadsheet

B <- read.csv('C:/Users/Carey Hedges/Desktop/Stats - Exam/question2.csv')
B

#All values that are measured need to correspond to the subject.  The data is grouped by the subject identification to ensure that all values for one individual are analysed as such.

B1 <- group_by(B, id)
B1

#We are not concerned with gender in this instance, as such, we exclude sex from the analysis.

B2 <- B1[,-2]
B2

#Tabulate data in order to run tests

B3 <- xtabs(handedness~first_stumble + final_position, data = B2)
B3

plot(B3, 
     main="Handedness and Line Position and First Stumble", 
     xlab = "First Stumble",
     ylab = "Final Postion", 
     col = c("Orange", "Red"))

#Table is a 2X2 table and the Fisher's test gives the exact p-value.

fisher.test(B3)

```

Results indicate that there is no link between handedness and markers of ataxic walking in intoxicated subjects (p=0.3002, CI(0.5576, 5.4757)).

#Question 3 - Amateur Runner

The data is numerical interval data.

The hypotheses are: 

Ho: The runner cannot predict her running time over her standard distance.
H1: The runner can presict her running time over a standard distance.

A test for correlation and linear regression will be used to evaluate whether or not there is good predictive value in the measurements that she currently has. 

The assumptions underlying the linear regression test are: 

1) A trend exists between running time and calories consumed.
2) Observations are independent of one another.
3) Measurements of the independent variable have been made correctly.
4) Residuals are normally distributed and are homoskedastic (same variance).

Assumptions underlying Pearson's correlation analyses (as used here) are:

1) Data should be interval or ratio data.
2) Data should have a linear relationship.
3) Data should not have any outliers or pivot points.
4) Variables should both be approximately normally distributed.

```{r question 3, echo = TRUE, fig.height=8, fig.width=8, fig.align='center', fig.path='./figures/', dev = "png"}



#Tell R to read the excel spread sheet

C <- read.csv('C:/Users/Carey Hedges/Desktop/Stats - Exam/question3.csv')
C

#Get an holistic overview of the data

summary(C)
head(C)
tail(C)

plot(C$time, C$calories, 
     main = "Calories Consumed versus Time Run", 
     ylab = "Calories Consumed (cal)",
     xlab = "Time (s)")

#Visual analysis suggests a possible linear regression model exists

#Remove the run column

C1 <- C[,-1]
C1

#Evaulate if there is a linear model that exists.

LMC <- lm(C1)
summary(LMC)


#A linear regression is present, perhaps a correlation exists.  I have used a Pearson's test as there are no outliers or influence points identified in the data set

cor.test(C$time, C$calories, method = "pearson")

#Check that the underlying assumptions of the tests have not been violated

#Residuals Check

qqnorm(LMC$residuals)
qqline(LMC$residuals)

#Hetereoskedacity check

plot(x = LMC$fitted, y = LMC$residuals,
     main = "Standard Deviation Across Measures",
     xlab = 'Fitted Values', 
     ylab = 'Residual Values')
abline(h=0, col = "gray")


```

Data indicates that the runner can predict the number of calories that she burns if she looks at her running time.  Running time and calories burned are inversely correlated (r=-0.6749, p=0.001522) and have a fairly strong negative association.  A linear model of the data is given by b|,-1.6252 t(17)=-3.772, p=0.001522 and R^2^= 0.4556, F(1,17)=14.22, p=0.001522.

This indicates that the runner could predict her calories that she burns when the time she runs falls between 1879 and 2174 seconds.  

Extrapolation of data in this model is bad methodology.  If the runner were to predict her calories consumed in 30 minutes this would correlate to 1800 seconds that lies outside this model and reduces that models ability to accurately predict her calorie loss over that time. However, crudely, extrapolation would indicate that caloric consumption is going to be close to or greater than  466cal and is extrapolated by this model to be 510.3379 calories ((-1.6252*y)+2629.4011=1800).

#Question 4 - Publishable Graphs

```{r r question 4, echo = TRUE, fig.height=8, fig.width=8, fig.align='center', fig.path='./figures/', dev = 'pdf'}

#Copy pasted data from word document.

foo <- rnorm(10000, mean = 60, sd = 3) # final mark
bar <- rnorm(10000, mean = 68, sd = 5) # project mark
baz <- rep(seq(from = 1997, to = 2006), each = 1) # years

year <- sample(baz, 150, replace = TRUE,
               prob = c(0.05, 0.05, 0.08, 0.08, 
                          0.1, 0.13, 0.14, 0.13, 0.12, 0.12))
project_mark <- sample(bar, 150, replace = TRUE)
final_mark <- sample(foo, 150, replace = TRUE)

plot_data <- data_frame(year = year,
                        project_mark = round(project_mark, 1),
                        final_mark = round(final_mark, 1)) %>%arrange(year)

#Visualise the data briefly.  As we are only assessing the 

summary(foo)
summary(bar)

#Plotted data adding a title, x and y axis labels and different limits for more space on the graph.

plot(bar, foo,
    main = "Relationship between Final Marks and Project Marks",
    xlab = "Project Mark (%)",
    ylab = "Final Mark (%)", 
    ylim = (c(45, 80)), 
    xlim = (c(45, 90)))

#r and p values should always be quoted on the graph.  I have run a correlation test to obtain an r value to place on the graph.

cor.test(bar, foo, method = "pearson")

plot(project_mark, final_mark,
    main = "Relationship between Final Marks and Project Marks",
    xlab = "Project Mark (%)",
    ylab = "Final Mark (%)", 
    ylim = (c(45, 80)), 
    xlim = (c(45, 90)),
    col=c(1997:2006))
legend ('topright', 
        inset = 0.05, 
        legend = c(1997:2006), 
        fill = c(1997:2006), 
        title = "Year")
text(50, 75, labels = 'r = - 0.05766')

```

There is a large density of points ontop of one another.  Extending the axes on to include zero and 100 make it very difficult to see what is happening in the figure.  Thus, I have shifted the default settings to incorperate a small amount of space around the points plotted on the axes.  